---
title: "Week 4 - Exercises"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(tidyverse) # require gives more informative errors than library
```


These exercises come from data carpentry. For each, show the resulting first couple lines of output using the head() command. If you were to print out each result, some of the tables would be pages long. See here under step 3:

https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf

This is really helpful where you can see the syntax and what it becomes in the knitted document.


# Exercise 0.

Here, include a 1page cheat sheet/summary of commands/functions for your reference. After you knit the document, ensure 1-page. Use the syntax as shown above (#, ##, ###, > , *, etc.). 


# Exercise 1 - summarize data

Let's look at the shrub-volume-data set again. Using group_by and summarize, find the maximum height of the shrubs by site and show result. Show both step by step and then translate to the pipe approach. You first need to add the readr and dplyr library (e.g., library(dplyr)) and then load in the data (function read.csv from readr package).

```{r}

# shift-control-m is the shortcut key for the pipe command
```

What is the average height for the shrubs in experiment 1?

```{r}

```

(answer: 4.7; need to show your work)

## Exercise 2 - Joining Tables

We have 2 additional datasets - sites and experiments. The sites data contains the lat/long (location data), and experiments contains information related to each experiment type.

Use inner_join to add both datasets to shrub_dims. You'll need to repeat inner_join for each.

```{r}



```

Your new table is 12 observations by 9 variables.


## Exercise 3

Using the surveys.csv dataset in your workspace:

1. Provide a new table that summarizes the number of individuals in each species ID (hint: group_by + summarize)

```{r}


```


2.  Provide a new table that summarizes the number of individuals in each species ID in each year (hint: group_by + summarize, final table is 535x3)

```{r}



  %>% head() # don't forget to NOT print the
             # whole thing out 
```

3. Find the mean mass of species DO in each year using filter, group_by, and summarize. If the data contains NA, omit so you have a value for each year.
 
```{r}


``` 

 [output for each of the above is here: https://datacarpentry.org/semester-biology/solutions/Portal-data-aggregation-R.txt]
 
 
## 4. Fix the following code 
 
The following code (which uses the piping command from last time) is supposed to import the shrub volume data and calculate the average shrub volume for each site and, separately, for each experiment, but it doesn't work!  

```{r, eval = FALSE}
read.csv("shrub-volume-data.csv")
shrub_data %>%
 mutate(volume = length * width * height) %>%
 group_by(site) %>%
 summarize(mean_volume = max(volume))
shrub_data %>%
 mutate(volume = length * width * height)
 group_by(experiment) %>%
 summarize(mean_volume = mean(volume))
```

Fix the code to complete the task. Remove the comment characters from all lines by selecting the whole chunk below and using "Comment/Uncomment Lines" from Code in the menu bar. The output should be the following: https://datacarpentry.org/semester-biology/solutions/Dplyr-fix-the-code-R.txt

```{r}
#read.csv("shrub-volume-data.csv")
#shrub_data %>%
#  mutate(volume = length * width * height) %>%
#  group_by(site) %>%
#  summarize(mean_volume = max(volume))
#shrub_data %>%
#  mutate(volume = length * width * height)
#  group_by(experiment) %>%
#  summarize(mean_volume = mean(volume))

```

## Exercise 5 - joins

Load surveys, plots, and species data into your workspace.

```{r}

```


* Use `inner_join()` to create a table that contains the information from both the surveys table and the species table.

```{r}



  %>% head()
```

* Use `inner_join()` twice to create a table that contains the information from all three tables.

```{r}



  %>% head()
```

* Use inner_join() and filter() to get a data frame with the information from the surveys and plots tables where the plot_type is Control.

```{r}



  %>% head()
```

[output should resemble the following: https://datacarpentry.org/semester-biology/solutions/Portal-data-joins-R.txt]

## 6 Comparisons

We want to do an analysis comparing the size of individuals on the Control plots to the Long-term Krat Exclosures using the plots, species, surveys datasets. Create a data frame with the year, genus, species, weight and plot_type for all cases where the plot type is either Control or Long-term Krat Exclosure. Only include cases where Taxa is Rodent. Remove any records where the weight is missing.

```{r}




```

## 7 Extract data

Using the Portal data surveys table (surveys.csv):

* Use $ to extract the weight column into a vector

```{r}




```

* Extract the hindfoot_length column into a vector and calculate the mean hindfoot length ignoring null values [hint: na.rm=TRUE].

```{r}





```

## Exercise 8

Suppose you have the following vectors:

```{r}
length <- c(2.2, 2.1, 2.7, 3.0, 3.1, 2.5, 1.9, 1.1, 3.5, 2.9)
width <- c(1.3, 2.2, 1.5, 4.5, 3.1, NA, 1.8, 0.5, 2.0, 2.7)
height <- c(9.6, 7.6, 2.2, 1.5, 4.0, 3.0, 4.5, 2.3, 7.5, 3.2)

```

Make a data frame that contains these three vectors as columns along with a genus column containing the name Taxus on all rows and a species column containing the word baccata on all rows.

```{r}



```